### 2022-03-04 paper/localpsf.tex
editing writing

In Section \ref{sec:intro} we noted that the integral kernel, $\Aker$, is inaccessible. Here we describe how we compute approximate kernel entries,
\begin{equation*}
\widetilde{\Phi}(y,x) \approx \Phi(y,x),
\end{equation*}
for arbitrary pairs of points $(y,x) \subset \Omega \times \Omega$.
% while Section \ref{sec:approximate_kernel_entries} we showed how we can cheaply evaluate approximate kernel entries at arbitrary pairs of points $(y,x) \in \Omega \times \Omega$.
To that end, we define the function
\begin{equation}
\label{eq:fxyxp}
f_{y,x}(x') := \frac{V(x)}{V(x')}\phi_{x_i}\left(y - \mu(x) + \mu(x')\right).
\end{equation}
It follows from inspection that 
\begin{equation*}
\Phi(y,x) = f_{y,x}(x),
\end{equation*}
and from the discussion in Section SEC, we expect that $\Phi(y,x) \approx f_{y,x}(x')$ when $x'$ is close to $x$. The factor $\frac{V(x)}{V(x_i)}$, which was not discussed in Section SEC, is a form of preconditioning that improves the quality of this approximation in situations where $V(x)$ varies rapidly. 
%, as is commonly the case for many operators of practical interest. In the Hessian for the Stokes inverse problem, for example, we observe that $V(x)$ varies over six orders of magnitude.
%As discussed in Section SECTION, we make the approximation
%\begin{equation*}
%	\Phi(y,x) = \phi_x(y) \approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right)
%\end{equation*}
%if $x'$ is close to $x$. Hence, w

We evaluate the approximate kernel entry $\widetilde{\Phi}(y,x)$ by performing radial basis function interpolation of the function $f_{y,x}(x')$ in the variable $x'$. The data used in the interpolation are the following (point,value) pairs
\begin{equation}
\left(x_i, f_{y,x}(x_i)\right), \qquad i=1,\dots,k,
\end{equation}
and $(x, f_{y,x}(x)) = (x,\Phi(y,x))$ is the (point,value) pair we approximate via interpolation.
Here $\{x_i\}_{i=1}^k \subset \Omega$ are the $k$ nearest sample points to $x$, excluding sample points for which $f_{y,x}(x_i)$ is not defined.
%a collection of sample points at which the impulse response $\phi_{x_i}$ is known.
%. Specifically, we interpolate $(\text{point}, \text{value})$ pairs $(x_i, f_i)$ at the point $x$, where $x_i$ are sample points, and $f_i$ are given as follows:
%\begin{equation}
%\label{eq:phi_mu_shift32}
%	f_i := \phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right).
%\end{equation}
%We must perform a new interpolation every time we evaluate $\Phi(y,x)$ at a new pair of points $(y,x)$, and constructing the H-matrix approximation of $\Phi$ requires evaluating $\Phi(y,x)$ at a large number of pairs of points $(y,x)$. 
%Interpolation using all sample points $x_i$ is too expensive, so we only use the $k$ nearest sample points to $x$, excluding sample points for which $f_{y,x}(x_i)$ is not defined. 
In numerical experiments we use $k=10$. The $k$ nearest sample points are found by querying a precomputed KD-tree \cite{kdtree} of all sample points.

@@

@@

%In order to compute $f_{y,x}(x_i)$, we must evaluate $\impulseresponse_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$, and we do this by evaluating $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$. Here $b$ is the index of the batch that contains the sample point $x_i$, and $\eta_b$ is the corresponding impulse response batch function that we have computed and stored.  
%Although it is typically recommended that one avoid evaluating finite element functions at arbitrary points [CITE], here we have no choice. 
To generate the right hand side vector $f$, we evaluate $f_{y,x}(x_i)$ at the points $x_i$ via the following process. First, we determine whether $y - \mu(x) + \mu(x_i) \in E_{x_i}$ by checking if
\begin{equation}
\label{eq:ellipsoid_check93}
(y - \mu(x))^T \Sigma^{-1}(x) (y - \mu(x))  \le \tau^2.
\end{equation}
If \eqref{eq:ellipsoid_check93} does not hold, then $y - \mu(x) + \mu(x_i) \notin E_{x_i}$ (i.e., $y - \mu(x) + \mu(x_i)$ is outside the estimated support of $\impulseresponse_{x_i}$), so we set $f_{y,x}(x_i)=0$ for the purpose of interpolation. If \eqref{eq:ellipsoid_check93} does hold, then $y - \mu(x) + \mu(x_i) \in E_{x_i}$. In this case, we look up the batch index $b$ for the batch of sample points that contains $x_i$, and evaluate
\begin{equation}
f_{y,x}(x_i) = \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right) = V(x)\eta_b\left(y - \mu(x) + \mu(x_i)\right).
\end{equation}
Recall from Section \ref{sec:get_impulse_response} that $\eta_b$ is the impulse response batch that we have computed and stored.
Evaluating $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$ requires some care, because the point $y - \mu(x) + \mu(x_i)$ is typically not a gridpoint of the mesh used to discretize the problem, even if $y$, $x$, and $x_i$ are gridpoints.
To evaluate $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$, we determine which mesh cell contains $y - \mu(x) + \mu(x_i)$, then we evaluate the relevant finite element basis functions on that mesh cell. Fortunately, the mesh cell containing $y - \mu(x) + \mu(x_i)$ was already determined as a side effect of querying the AABB tree of mesh cells when we checked whether $y - \mu(x) + \mu(x_i) \in \Omega$.

\subsubsection{Improvements based on symmetry}
\label{sec:symmetric_improvements}
If $\Aop$ is symmetric, impulse responses of $\Aop$ are also impulse responses of $\Aop^T$. Let $\{y_i\}_{i=1}^k$ be the $k$ nearest sample points to $y$, excluding points $y_i$ such that $x - \mu(y) + \mu(y_i) \notin \Omega$. We could form an alternative approximation of $\Phi(y,x)$ by interpolating the (point, value) pairs
\begin{equation}
\label{eq:phi_mu_shift_y}
\left(y_i,f_{x,y}(y_i)\right), \qquad i=1,\dots,k,
\end{equation}
in the same manner as before, but with $x$ and $y$ now taking opposite roles. We combine both of these interpolation schemes to improve the approximation accuracy at negligible additional. The two interpolation schemes are combined by placing function values 
\begin{equation*}
\left(\widehat{f}_x\right)_i := \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $x_i-x$, placing function values
\begin{equation*}
\left(\widehat{f}_y\right)_i := \frac{V(y)}{V(y_i)}\phi_{y_i}\left(x - \mu(y) + \mu(y_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $y_i-y$, and interpolating the combination of these $2k$ points and values at zero.

In detail, define 
\begin{equation*}
\widehat{x_i} := x_i - x \qquad \text{and} \qquad \widehat{y_i} := y_i - y.
\end{equation*}
and let
\begin{equation*}
\widehat{B} = \begin{bmatrix}
\widehat{B}_{xx} & \widehat{B}_{xy} \\
\widehat{B}_{yx} & \widehat{B}_{xx}
\end{bmatrix}, \qquad \widehat{w} = \begin{bmatrix}\widehat{w}_x \\ \widehat{w}_y\end{bmatrix}, \qquad \widehat{f} = \begin{bmatrix}
\widehat{f}_x \\ \widehat{f}_y
\end{bmatrix},
\end{equation*}
be a $2k \times 2k$ block matrix and $2k \times 1$ block vectors, respectively, with the block entries of $\widehat{B}$ be given by
\begin{align*}
\left(\widehat{B}_{xx}\right)_{ij} = b\left(\|\widehat{x}_i - \widehat{x}_j\|\right), &\qquad
\left(\widehat{B}_{xy}\right)_{ij} = b\left(\|\widehat{x}_i - \widehat{y}_j\|\right), \\
\left(\widehat{B}_{yx}\right)_{ij} = b\left(\|\widehat{y}_i - \widehat{x}_j\|\right), &\qquad \left(\widehat{B}_{yy}\right)_{ij} = b\left(\|\widehat{y}_i - \widehat{y}_j\|\right),
\end{align*}
for $i,j=1,\dots,k$. We approximate $\Phi(y,x)$ as follows:
\begin{equation}
\widetilde{\Phi}(y,x) := \sum_{i=1}^k \left(w_x\right)_i~ b\left(\|\widehat{x}_i\|\right) + \sum_{i=1}^k \left(w_y\right)_i~ b\left(\|\widehat{y}_i\|\right),
\end{equation}
where $\widehat{w}$ is the solution of the following regularized least-squares problem,
\begin{equation*}
\min_{\widehat{w}} \frac{1}{2}\|\widehat{B} \widehat{w} - \widehat{f}\|^2 + \frac{\alpha}{2}\|\widehat{w}\|^2,
\end{equation*}
for a small regularization parameter $\alpha$. 
We solve the regularized least squares problem by QR factorization of the block $4k \times 2k$ matrix $\begin{bmatrix}\widehat{B} & I\end{bmatrix}^T$. We regularize the problem because of numerical issues that arise if there are points $\widehat{x}_i$ and $\widehat{y}_j$ in the interpolation that are very close to each other, a situation that arises often in practice. Indeed, situations where $\widehat{x}_i = \widehat{y}_j$ are inevitable when a regular grid is used to discretize the problem. In this case, $\widehat{B}$ is (nearly) singular, so it is numerically ill-advised to solve the standard interpolation equation $\widehat{B} \widehat{w} = \widehat{f}$.

>>>>

In Section \ref{sec:intro} we noted that the integral kernel, $\Aker$, is inaccessible. Here we describe how we compute approximate kernel entries,
\begin{equation*}
\widetilde{\Phi}(y,x) \approx \Phi(y,x),
\end{equation*}
for arbitrary pairs of points $(y,x) \subset \Omega \times \Omega$ via interpolation.
% while Section \ref{sec:approximate_kernel_entries} we showed how we can cheaply evaluate approximate kernel entries at arbitrary pairs of points $(y,x) \in \Omega \times \Omega$.
To that end, we define the function
\begin{equation}
\label{eq:fxyxp}
f_{y,x}(x') := \frac{V(x)}{V(x')}\phi_{x_i}\left(y - \mu(x) + \mu(x')\right).
\end{equation}
It follows from inspection that 
\begin{equation*}
\Phi(y,x) = f_{y,x}(x),
\end{equation*}
and from the discussion in Section SEC, we expect that $\Phi(y,x) \approx f_{y,x}(x')$ when $x'$ is close to $x$. The factor $\frac{V(x)}{V(x_i)}$, which was not discussed in Section SEC, is a form of preconditioning that improves the quality of this approximation in situations where $V(x)$ varies rapidly. 
%, as is commonly the case for many operators of practical interest. In the Hessian for the Stokes inverse problem, for example, we observe that $V(x)$ varies over six orders of magnitude.
%As discussed in Section SECTION, we make the approximation
%\begin{equation*}
%	\Phi(y,x) = \phi_x(y) \approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right)
%\end{equation*}
%if $x'$ is close to $x$. Hence, w

We evaluate $\widetilde{\Phi}(y,x)$ by performing radial basis function interpolation of the function $f_{y,x}(x')$ in the variable $x'$. The data used in the interpolation are the following (point,value) pairs
\begin{equation}
\left(x_i, f_{y,x}(x_i)\right), \qquad i=1,\dots,k,
\end{equation}
and $(x, f_{y,x}(x)) = (x,\Phi(y,x))$ is the (point,value) pair we approximate via interpolation.
Here $\{x_i\}_{i=1}^k \subset \Omega$ are the $k$ nearest sample points to $x$, excluding sample points for which $f_{y,x}(x_i)$ is not defined.
%a collection of sample points at which the impulse response $\phi_{x_i}$ is known.
%. Specifically, we interpolate $(\text{point}, \text{value})$ pairs $(x_i, f_i)$ at the point $x$, where $x_i$ are sample points, and $f_i$ are given as follows:
%\begin{equation}
%\label{eq:phi_mu_shift32}
%	f_i := \phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right).
%\end{equation}
%We must perform a new interpolation every time we evaluate $\Phi(y,x)$ at a new pair of points $(y,x)$, and constructing the H-matrix approximation of $\Phi$ requires evaluating $\Phi(y,x)$ at a large number of pairs of points $(y,x)$. 
%Interpolation using all sample points $x_i$ is too expensive, so we only use the $k$ nearest sample points to $x$, excluding sample points for which $f_{y,x}(x_i)$ is not defined. 
In numerical experiments we use $k=10$. The $k$ nearest sample points are found by querying a precomputed KD-tree \cite{kdtree} of all sample points.

@@

\subsubsection{Radial basis function interpolation}

@@

%In order to compute $f_{y,x}(x_i)$, we must evaluate $\impulseresponse_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$, and we do this by evaluating $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$. Here $b$ is the index of the batch that contains the sample point $x_i$, and $\eta_b$ is the corresponding impulse response batch function that we have computed and stored.  
%Although it is typically recommended that one avoid evaluating finite element functions at arbitrary points [CITE], here we have no choice.

%\subsubsection{Evaluating $f_{y,x}(x_i)$}

To generate the right hand side vector $f$, we evaluate $f_{y,x}(x_i)$ at the points $x_i$ via the following process. First, we determine whether $y - \mu(x) + \mu(x_i) \in E_{x_i}$ by checking if
\begin{equation}
\label{eq:ellipsoid_check93}
(y - \mu(x))^T \Sigma^{-1}(x) (y - \mu(x))  \le \tau^2.
\end{equation}
If \eqref{eq:ellipsoid_check93} does not hold, then $y - \mu(x) + \mu(x_i) \notin E_{x_i}$ (i.e., $y - \mu(x) + \mu(x_i)$ is outside the estimated support of $\impulseresponse_{x_i}$), so we set $f_{y,x}(x_i)=0$ for the purpose of interpolation. If \eqref{eq:ellipsoid_check93} does hold, then $y - \mu(x) + \mu(x_i) \in E_{x_i}$. In this case, we look up the batch index $b$ for the batch of sample points that contains $x_i$, and evaluate
\begin{equation}
f_{y,x}(x_i) = \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right) = V(x)\eta_b\left(y - \mu(x) + \mu(x_i)\right).
\end{equation}
%Recall from Section \ref{sec:get_impulse_response} that $\eta_b$ is the impulse response batch that we have computed and stored.
Evaluating the impulse response batch $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$ requires some care, because the point $y - \mu(x) + \mu(x_i)$ is generally not a gridpoint of the mesh used to discretize the problem, even if $y$, $x$, and $x_i$ are gridpoints.
To evaluate $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$, we determine which mesh cell contains $y - \mu(x) + \mu(x_i)$, then we evaluate the relevant finite element basis functions on that mesh cell. Fortunately, the mesh cell containing $y - \mu(x) + \mu(x_i)$ was already determined as a side effect of querying the AABB tree of mesh cells when we checked whether $y - \mu(x) + \mu(x_i) \in \Omega$.

\subsubsection{Improved interpolation for symmetric operators}
\label{sec:symmetric_improvements}
If $\Aop$ is symmetric, impulse responses of $\Aop$ are also impulse responses of $\Aop^T$. Let $\{y_i\}_{i=1}^k$ be the $k$ nearest sample points to $y$, excluding points $y_i$ such that $x - \mu(y) + \mu(y_i) \notin \Omega$. We could form an alternative approximation of $\Phi(y,x)$ by interpolating the (point, value) pairs
\begin{equation}
\label{eq:phi_mu_shift_y}
\left(y_i,f_{x,y}(y_i)\right), \qquad i=1,\dots,k,
\end{equation}
in the same manner as before, but with $x$ and $y$ now taking opposite roles. We combine both of these interpolation schemes to improve the approximation accuracy at negligible additional. The two interpolation schemes are combined by placing function values 
\begin{equation*}
\left(\widehat{f}_x\right)_i := \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $x_i-x$, placing function values
\begin{equation*}
\left(\widehat{f}_y\right)_i := \frac{V(y)}{V(y_i)}\phi_{y_i}\left(x - \mu(y) + \mu(y_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $y_i-y$, and interpolating the combination of these $2k$ points and values at zero.

In detail, define 
\begin{equation*}
\widehat{x_i} := x_i - x \qquad \text{and} \qquad \widehat{y_i} := y_i - y.
\end{equation*}
and let
\begin{equation*}
\widehat{B} = \begin{bmatrix}
\widehat{B}_{xx} & \widehat{B}_{xy} \\
\widehat{B}_{yx} & \widehat{B}_{xx}
\end{bmatrix}, \qquad \widehat{w} = \begin{bmatrix}\widehat{w}_x \\ \widehat{w}_y\end{bmatrix}, \qquad \widehat{f} = \begin{bmatrix}
\widehat{f}_x \\ \widehat{f}_y
\end{bmatrix},
\end{equation*}
be a $2k \times 2k$ block matrix and $2k \times 1$ block vectors, respectively, with the block entries of $\widehat{B}$ be given by
\begin{align*}
\left(\widehat{B}_{xx}\right)_{ij} = b\left(\|\widehat{x}_i - \widehat{x}_j\|\right), &\qquad
\left(\widehat{B}_{xy}\right)_{ij} = b\left(\|\widehat{x}_i - \widehat{y}_j\|\right), \\
\left(\widehat{B}_{yx}\right)_{ij} = b\left(\|\widehat{y}_i - \widehat{x}_j\|\right), &\qquad \left(\widehat{B}_{yy}\right)_{ij} = b\left(\|\widehat{y}_i - \widehat{y}_j\|\right),
\end{align*}
for $i,j=1,\dots,k$. We approximate $\Phi(y,x)$ as follows:
\begin{equation}
\widetilde{\Phi}(y,x) := \sum_{i=1}^k \left(w_x\right)_i~ b\left(\|\widehat{x}_i\|\right) + \sum_{i=1}^k \left(w_y\right)_i~ b\left(\|\widehat{y}_i\|\right),
\end{equation}
where $\widehat{w}$ is the solution of the following regularized least-squares problem,
\begin{equation*}
\min_{\widehat{w}} \frac{1}{2}\|\widehat{B} \widehat{w} - \widehat{f}\|^2 + \frac{\alpha}{2}\|\widehat{w}\|^2,
\end{equation*}
for a small regularization parameter $\alpha$. 
We solve the regularized least squares problem by QR factorization of the block $4k \times 2k$ matrix $\begin{bmatrix}\widehat{B} & I\end{bmatrix}^T$. We regularize the problem because of numerical issues that arise if there are points $\widehat{x}_i$ and $\widehat{y}_j$ in the interpolation that are very close to each other, a situation that arises often in practice. Indeed, situations where $\widehat{x}_i = \widehat{y}_j$ are inevitable when a regular grid is used to discretize the problem. In this case, $\widehat{B}$ is (nearly) singular, so it is numerically ill-advised to solve the standard (un-regularized) interpolation equation $\widehat{B} \widehat{w} = \widehat{f}$.
